A framework for pipeline benchmarking and its application to single-cell RNAseq clustering

### 2020-01-29 bmc_article.tex
Updates from Overleaf

\begin{abstract} % abstract
%\parttitle{Background} %if any
We present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of single-cell RNAseq analysis pipelines using datasets with known cell identities. We discuss common steps of such analyses, including filtering and doublet detection (offering a new package), normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete key recommendations about analysis choices.
\end{abstract}

@@

Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization. This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}. Data processing is a critical step in any scRNAseq analysis as these data are affected by various technical effects (batch effect, dropout, ...) that have to be corrected to distinguish the biological variation form the technical variation. The choice of the scRNAseq protocol is already of prime importance as different protocols differ greatly in view of parameters such as gene detection or mappability \cite{MereuCellAtlas2019}, which can in turn affect downstream analysis. In view of the intrinsic challenges of scRNAseq technology and the growing number of tools and studies complexity, it is critical to evaluate the available tools to leverage future studies as it can ultimately increase the power of further differential expression analysis \cite{viethSystematic2019}.

@@

Here, we develop a flexible R framework for pipeline comparison and evaluate in details the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We harness real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} (\cite{CrowellMuscat2019})). 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \texttt{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \texttt{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations on the fly to avoid storing potentially large intermediate data. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets produced by \citep{tianMixology2018} with cells of different genetic identity, and for which therefore SNP information can be used as ground truth. We tested \texttt{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \texttt{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \texttt{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification. \texttt{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \texttt{scran} and \texttt{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets.

In addition to these methods, we reasoned that an approach such as \texttt{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets (i.e. doublets from different cell types) more efficiently. We therefore developed a simple and fast Biocondcutor package implementing this for doublet detection, \texttt{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A). \texttt{scDblFinder} achieved comparable or better accuracy than top alternatives while running in a very reasonable time (Figure \ref{fig:figure3}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C. We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by scDblFinder as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \texttt{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, but not always: for instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In addition, in our experience 10X datasets also exhibit a very tight correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

@@

We next investigated the impact of filtering according to various criteria (see methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g. library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \citep{PapadimitriouHu1998}, and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias. Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties, such as high library size, rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see methods).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \texttt{scater} (see methods for details).}
    \label{fig:figure4}
\end{figure}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \texttt{Seurat}, we tested \texttt{scran}'s pooling-based normalization \citep{lunPooling2016}, \texttt{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard \texttt{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\texttt{sctransform}'s variance-stabilizing transformation involves something analogous to scaling). 
\texttt{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq\citep{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development \texttt{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \texttt{Seurat} normalization resulted in fairly high correlation with technical covariates. \texttt{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling. The only exception is \texttt{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.

We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width and mutual information. We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \texttt{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with Smart-seq protocol (Koh and Kumar datasets).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures. `Seurat.feat_mt_regress' indicates the regressing out of the number of features and proportion of mitochondrial reads during scaling, represents, and `Seurat.feat_regress` of the number of features only. A:} Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy, measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g. \texttt{sctransform}) or lower (e.g. stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).

@@

A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq. \texttt{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions (>3) rely on a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types. Another common practice in features selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \texttt{sctransform} provides its own ordering of genes based on transformed variance.

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed whether a method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives. Analysis of variance performed on a standard \texttt{Seurat} normalization or on \texttt{sctransform} data were highly correlated (Supplementary Figure 11A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight the differences between the methods, although a higher number of selected genes decreased the differences between the methods (Supplementary Figures 12-14). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding to the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but surprisingly did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterix denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \texttt{Seurat} normalization (left) or \texttt{sctransform} (right). The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

We next evaluated how the use of the different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/ deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).

@@

For research articles this section should discuss the implications of the findings in context of existing research and highlight limitations of the study. For study protocols and methodology manuscripts this section should include a discussion of any practical or operational issues involved in performing the study and any issues not covered in other sections.

>>>>

\begin{abstract} % abstract
%\parttitle{Background} %if any
The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis is currently only insufficiently accompanied by proper and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of key concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool and permit further extending benchmarks (\url{https://github.com/plger/pipeComp}).

\end{abstract}

@@

Single-cell RNA-sequencing (scRNAseq) and the set of attached analysis methods are evolving fast, with more than 560 software tools available to the community \cite{ZappiaDB2018}, roughly half of which are dedicated to tasks related to data processing such as clustering, ordering, dimension reduction or normalization. This increase in the number of available tools follows the development of new sequencing technologies and the growing number of reported cells, genes and cell populations \cite{SvenssonDB2019}. Data processing is a critical step in any scRNAseq analysis, affecting downstream analysis and interpretation. It is therefore critical to evaluate the available tools to leverage future studies as it can ultimately increase the power of further differential expression analysis \cite{viethSystematic2019}.

@@

Here, we develop a flexible R framework for pipeline comparison and evaluate in details the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We harnessed real datasets of known cell composition (Table \ref{tab:table1}) and develop new evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and \ref{fig:figure1} for a description of the datasets. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}). 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compare not only competing approaches, but also more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (\ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. The \textit{CellBench} package was recently proposed to address a similar need\citep{su_cellbench}, offering an elegant piping syntax to combine alternative methods at successive steps. In addition to offering the same capabilities, a key additional feature of \textit{pipeComp} is that, since the evaluation metrics are not treated as a step but parallel to it, they can be performed at various points in the pipeline without needing to store the potentially large intermediate data, and thereby enabling combinatorial benchmarks. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect. For example, by far the most important determinant of the ARI is the number of detected clusters: the farther it is from the actual number of subpopulations, the worse the ARI. In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline.

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We therefore first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, and for which therefore SNP information can be used as ground truth. We included a further such dataset\citep{kangMultiplexedDemuxlet2018} not included for the rest of our benchmark due to the absence of true cell labels beside SNPs. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types) and can be identified through other means.

We tested \textit{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification. \textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets. In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently. We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see methods).

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while running in a very reasonable time (Figure \ref{fig:figure3}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C. We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function of \textit{scater} that measures, for a given control property, the number of median absolute deviations of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For instance, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

@@

We next investigated the impact of filtering according to various criteria (see methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g. library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of median absolute deviations (MADs) from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \citep{PapadimitriouHu1998}, and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias. Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see methods).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see methods for details).}
    \label{fig:figure4}
\end{figure}

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \citep{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves something analogous to scaling). 
\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq\citep{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates. \textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling. The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.

We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width\citep{RousseeuwSil1987} and mutual information. We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with Smart-seq protocol (Koh and Kumar datasets).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' indicates the regressing out of the number of features and proportion of mitochondrial reads during scaling, represents, and `Seurat.feat\_regress` of the number of features only.  Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy, measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g. \textit{sctransform}) or lower (e.g. stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).

@@

A standard clustering pipeline typically involves a step of highly-variable genes selection, which is complicated by the digital nature and the mean-variance relationship of (sc)RNAseq. \textit{Seurat}'s earlier approaches involved the use of dispersion estimates standardized for the mean expression levels, while more recent versions (>3) rely on a different measure of variance, again standardized. Of note, while adjusting for the mean-variance relationship removes much of the bias towards highly-expressed genes, it is plausible that this relationship may in fact sometimes reflects biological relevance and would be helpful in classifying cell types. Another common practice in features selection is to use those with the highest mean expression, and recently, \citep{townesGlmpca2019} instead suggested to use deviance, while \textit{sctransform} provides its own ordering of genes based on transformed variance.

Reasoning that a selection method should ideally select genes whose variability is more between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three such alternatives. Analysis of variance performed on a standard \textit{Seurat} normalization or on \textit{sctransform} data were highly correlated (Supplementary Figure 11A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding to the percentage of deviance explained however, the standardized measures were often superior(Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good), but did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterix denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right). The methods `vst.varExp' and `devianceExplained' are the estimates used in \textbf{A} to evaluate the selection methods, and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/ deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).

@@
